---
title: "Persisting somatic symptoms following COVID-19 as well as mental health status are associated with changes in neurotransmitter precursor amino acid levels – a psychoneuroimmunological study"
subtitle: "Supplementary Material"
author: "PSY Team"
date: "`r format(Sys.time(), '%Y-%m-%d')`"

output: 
  bookdown::word_document2:   
    reference_docx: ms_template.docx
    
bibliography: infl_biblio.bib
csl: frontiers-in-immunology.csl
header-includes:
  \usepackage{longtable}
  \usepackage{tabu}
  \usepackage{caption}
  \usepackage{makecell}
  \usepackage{pdflscape}
  \usepackage{array}
  \usepackage{booktabs}
  \usepackage{threeparttable}
  \usepackage{threeparttablex}
  \usepackage{wrapfig}
  \usepackage{multirow}
  \usepackage[normalem]{ulem}
  \usepackage{colortbl}
  \usepackage{xcolor}
  \usepackage{float} \floatplacement{figure}{H} \floatplacement{table}{H}
  \newcommand{\beginsupplement}{\setcounter{table}{0}  \renewcommand{\thetable}{S\arabic{table}}       \setcounter{figure}{0} \renewcommand{\thefigure}{S\arabic{figure}}}
---

```{r, setup, include = FALSE}

knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, 
                      dev = "png", dpi = 600, 
                      tab.cap.pre = 'Supplementary Table ', tab.cap.sep = " ", 
                      tab.cap.style = 'Table Heading')

set_flextable_defaults(font.family = 'Cambria', font.size = 10)


```

\newpage

# Supplementary Methods

## External data import and transformation

Proteome and metabolome data in form of normalized, age- and sex-adjusted, log~2~-transformed serum levels as well as clinical information (sex, SARS-CoV-2 infection status, COVID-19 severity, timepoint, post-COVID-19 syndrome status and particular symptoms) for the INCOV cohort were extracted from supplementary tables accompanying the report by Su and colleagues [@Su2022]. 

Characteristic of the INCOV cohorts is presented in __Table 2__. 
Numbers of available INCOV cohort samples and the sampling timepoints are shown in __Supplementary Table S\@ref(tab:tab-samples)__.

## Software

Proteome and metabolome data were analyzed with R version 4.2.0. 
General data transformation tasks were accomplished with the _tidyverse_ package bundle [@Wickham2019], _rlang_ [@Henry2022] and the development package _trafo_ (https://github.com/PiotrTymoszuk/trafo). 
Statistical data analysis was done with the packages _rstatix_ [@Kassambara2021], _ggpubr_ [@Kassambara2020] along with the development package _ExDA_ (https://github.com/PiotrTymoszuk/ExDA).
Results were visualized with tools provided by the packages _ggplot2_ [@Wickham2016], _cowplot_ [@Wilke2019] and _ExDA_. 
Manuscript and supplementary tables were created with _flextable_ [@Gohel2022]. 
Supplementary Material file was written in the _rmarkdown_ environment [@Allaire2022] and rendered with the _knitr_ [@Xie2022] and _booksdown_ [@Xie2016] packages.

## Statistical analysis of the cytokine and metabolite INCOV cohort data

Distribution normality and homogeneity of normalized cytokine and metabolite serum levels was assessed by Shapiro-Wilk and Levene tests, respectively. 
Comparison of normalized serum cytokine and metabolite levels between healthy controls and COVID-19 individuals at consecutive timepoints after symptom onset was done with Kruskal-Wallis test. 
Differences between healthy controls and consecutive timepoints of COVID-19 were investigated with Mann-Whitney U test corrected for multiple testing with Benjamini-Hochberg method [@Benjamini1995]. 
Correlation of serum cytokine and metabolite levels in healthy controls and COVID-19 individuals at consecutive timepoints was analyzed with Spearman test.
For comparison of normalized serum levels of cytokines and metabolites during COVID-19 convalescence between healthy controls, COVID-19 individuals without persistent symptoms, COVID-19 individuals with persistent putative psychiatric symptoms (depression, anxiety or sleep disorders) and COVID-19 individuals with other persistent symptoms Kruskal-Wallis test with Mann-Whitney post-hoc test was employed.

## Data and code availability

Anonymized local cohort data will be made available upon request to the corresponding author. 
The INCOV cohort data are publicly available.
The study pipeline code is available at https://github.com/PiotrTymoszuk/stigma_validation.

\newpage

# Supplementary Tables

```{r tab-samples, tab.cap = 'Number of available samples and sampling timepoints in the INCOV cohort.'}

flextable::flextable(suppl_tabs$samples) %>% 
  width(1, width = 4, unit = 'cm') %>% 
  width(2, width = 6, unit = 'cm') %>% 
  width(3, width = 4, unit = 'cm') %>% 
  footnote(1, 2, value = as_paragraph('Median with interquartile range, days after symptom onset.'), ref_symbols = 'a', part = 'header') %>% 
  theme_vanilla

```

\newpage

# Supplementary Figures

```{r fig-incov-neuro-cyto, fig.width = mm_inch(suppl$incov_neuro_cyto$w), fig.height = mm_inch(suppl$incov_neuro_cyto$h), fig.cap = 'Serum levels of cytokines in healthy controls, complete COVID-19 recovery, non-neurological and neurological persistent symptoms in the INCOV cohort.'}

suppl$incov_neuro_cyto$plot

```

__Figure \@ref(fig:fig-incov-neuro-cyto). Serum levels of cytokines in healthy controls, complete COVID-19 recovery, non-neurological and neurological persistent symptoms in the INCOV cohort.__ 
_Normalized serum levels of IL6, IL10, TNF-alpha (TNF), IFN-gamma (IFNG) were extracted from the INCOV study data for healthy controls and COVID-19 participants during convalescence (median: 64 days after symptom onset). The serum levels were compared between healthy controls, COVID-19 subjects without persistent symptoms, COVID-19 subjects with putative neurological persistent symptoms (self-reported neurological symptoms, blurry vision, dizziness, feet or hand pain, smell or taste disorders) and COVID-19 subjects with other non-neurological persistent symptoms with Kruskal-Wallis test. Pairwise comparisons between the groups were done with Benjamini-Hochberg-corrected Mann-Whitney post-hoc test. Each point in the plots represent a single sample, boxes depict medians with interquartile ranges (IQR), whiskers span over 150% IQR. Kruskal-Wallis test p values are displayed in the plot captions. Significant and near-significant (p < 0.1) post-hoc test results are indicated in the plots. Numbers of complete observations are shown below the plots._

\newpage

```{r fig-incov-neuro-trp, fig.width = mm_inch(suppl$incov_neuro_trp$w), fig.height = mm_inch(suppl$incov_neuro_trp$h), fig.cap = 'Serum levels of TRP degradation products in healthy controls, complete COVID-19 recovery, non-neurological and neurological persistent symptoms in the INCOV cohort.'}

suppl$incov_neuro_trp$plot

```

__Figure \@ref(fig:fig-incov-neuro-trp). Serum levels of TRP degradation products in healthy controls, complete COVID-19 recovery, non-neurological and neurological persistent symptoms in the INCOV cohort.__ 
_Normalized serum levels of tryptophan (TRP), serotonin (5-HT), kynurenine (KYN) and quinolinic acid (QUIN) were extracted from the INCOV study data for healthy controls and COVID-19 participants during convalescence (median: 64 days after symptom onset). The serum levels were compared between healthy controls, COVID-19 subjects without persistent symptoms, COVID-19 subjects with putative neurological persistent symptoms (self-reported neurological symptoms, blurry vision, dizziness, feet or hand pain, smell or taste disorders) and COVID-19 subjects with other non-neurological persistent symptoms with Kruskal-Wallis test. Pairwise comparisons between the groups were done with Benjamini-Hochberg-corrected Mann-Whitney post-hoc test. Each point in the plots represent a single sample, boxes depict medians with interquartile ranges (IQR), whiskers span over 150% IQR. Kruskal-Wallis test p values are displayed in the plot captions. Significant and near-significant (p < 0.1) post-hoc test results are indicated in the plots. Numbers of complete observations are shown below the plots._

\newpage

# References
